Data Migration under a Changing Schema in Ampersand

### 2023-10-08 2022Migration/articleMigrationFACS.tex
WIP

\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be upgraded regularly to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly upgrades of existing systems.
   In upgrades, schema changes cannot always be avoided, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   The risk and effort of such data migrations explains why development teams try to avoid schema changes at all costs.

   Data migration for other purposes than schema change has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers can and will avoid schema changes and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
   Another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014}, defines a migration as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   This does not cater for schema changes, however.
   Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
   If the schema is not available at compile time, the work will have to be done at runtime.
   This requires versioned storage of production data and an overhead in performance.
   This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
   The prototypical use case for that is to frequently release upgrades of information systems in production,
   where the semantic integrity of data must be preserved across schema changes.
   Another use case is application integration for multiple dispersed data sources with explicit schemas.

@@

   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, invariants that change yield extra work for software engineers to write the transform code,
   for which ETL tools typically provide little support.
   The pressure to decrease deploy times calls for further automation of SCDMs.
   Hence, we can expect more frequent SCDMs, so downtime for the sake of migration becomes less acceptable.
   That is why our research aims for zero downtime.

   Another practical problem is that of data quality.
   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean up data.
   Some of that must be done before the migration; some can wait till after the migration.
   We can capture part of the data quality problem as a requirement to satisfy semantic constraints.
   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
   a registration of valid addresses can be checked automatically.
   In a formalism like Ampersand, which allows us to express such constraints, we can add constraints for data quality to the schema.
   This allows us to signal the data pollution at runtime.
   Some forms of data pollution cannot be detected in this way, however.
   An example is when a person has deliberately specified a false name without violating any conceivable constraint.

   To make the case for zero downtime, we must distinguish between two types of constraint:
\begin{enumerate}
\item Invariant\\
   An \define{invariant} is a constraint that is always true in a system.
   This corresponds to the classical database transaction.
   Violation of an invariant means it has to be restored without the outside world noticing.
   This requires either human intervention or an automated procedure.
   Once it is restored, the system commits to the new state, which satisfies the constraint.
   If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
   In both cases, the invariant remains true.

   A special case of an invariant is the \define{enforcement rule}.
   As soon as an enforce rule is violated, the system restores it automatically, without human intervention.
\item Business constraint\\
   A \define{business constraint} is a constraint that can be violated temporarily until a user restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   We consider business constraints not to be invariants because users can violate them.
   They are not true all the time.
\end{enumerate}
   Suppose some invariant, $u$, of the {\em desired system} does not apply to the {\em existing system},
   and suppose it would be violated if applied to the data of the existing system.
   Now, if we try to spin up the desired system on the data set of the existing system, we have a problem.
   Every invariant of the desired system must be satisfied but $u$ isn't.
   So we cannot spin it up.
   In many cases, clearing out all violations before spinning up the desired system might take too much time and new violations will arise in the meantime.
   To solve this problem, we implement every invariant in the migration system initially as a business constraint,
   letting users restore invariance while the new system is running.
   For this purpose, we must define an intermediate system, the \define{migration system},
   which implements the invariants of the desired system as business constraints.
   This migration system must ensure that every violation that is fixed will not return as a violation.
   In this way, the business constraint gradually turns into an invariant.
   Since the number of violations is finite, the business can resolve these violations in finite time.
   New violations that occur are under control of the desired system,
   so the business experiences the migration system as the desired system.
   In this way, the migration system bridges the gap and users get a zero downtime SCDM.

@@

\section{Analysis}
\label{sct:Analysis}
\subsection{Information Systems}
   This section defines information systems, so we can formally define the migration procedure from one system to another.

   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
   The data in the system constitutes the collective memory of the business,
   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
   %This paper uses semantic constraints to represent this meaning.

   Actors (both users and computers) are changing the data in a system continually.
   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
   Events that the system detects may cause the state to change.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
   \end{center}
\caption{Anatomy of an information system}
\label{fig:pre-migration}
\end{figure}
   To keep our theory technology independent, we assume data sets to contain triples.
   This makes our theory valid for any kind of database that triples can represent,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
   An important aspect of the system's semantics are the invariants, sometimes known as integrity rules.
   %The purpose of these rules is to keep all constraints satisfied.

@@

   Before the migration, the existing system satisfies its invariants (by definition).
   When the migration system compiles correctly, it satisfies its own invariants,
   so it can be deployed in production.
   That is the MoT.
   The task of the migration system is to satisfy the invariants of the desired system
   so that it can be deployed in production.
   For some of these invariants, enforcement rules can satisfy them automatically.
   There may be invariants, however, that require human interaction to resolve.
   Once all invariants of the desired system are satisfied, the migration is ready for the MoC.
   So, the schema of the migration system must contain the rules to satisfy all invariants of the desired system.

@@

\subsection{Violations}
\label{sct:Violations}
   To satisfy a constraint $u$ means that it is true for all triples in the data set.
   Let us assume that we can represent each violation of a constraint $u$ as a pair
   and let the function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produce these violations.
   Note that this assumption implies that rules have a signature, just like a relations.
   So, we can represent an invariant $u$ as a triple $\triple{\violC{u}}{A}{B}$
   that satisfies:
\begin{equation}
   \pair{a}{b}\in\viol{\triple{\violC{u}}{A}{B}}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
\end{equation}
   We represent an enforce rule $u$ as a pair $\pair{\declare{n}{A}{B}}{\violC{u}}$,
   where relation $\declare{n}{A}{B}$ is used to restore invariance of the enforce rule.
   The enforce rule satisfies:
\begin{equation}
   \pair{a}{b}\in\viol{u}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B\ \wedge\ \viol{u}{\dataset\cup\triple{a}{\declare{n}{A}{B}}{b}}=\emptyset
\end{equation}

\subsection{Schemas}
\label{sct:Schemas}
   Schemas serve to capture the semantics of a system~\cite{Spivak2012}.
   It defines concepts, relations, and invariants.
   We assume that a software engineer defines a schema on design time, and a compiler checks whether the semantics are consistent.
   Error messages from the compiler are prohibitive for generating code,
   which prevents a substantial class of mistakes to ever reach end users.

@@

   Each invariant in a schema serves to constrain the data set at runtime, to ensure its semantic integrity.
   Every invariant is an element of an infinite set called $\Rules$.
   In this paper, variables $u$ and $v$ represent rules.
   A invariant $u$ in the schema is a triple $\triple{\violC{u}}{A}{B}$ such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
   In Ampersand, rules are well-typed, just like relations:
   Every violation $\pair{a}{b}$ of a invariant $\triple{\violC{u}}{A}{B}$ must satisfy $a\instance A$ and $b\instance B$.
   % For every data set $\dataset$, the set $\viol{u}{\dataset}$ is a subset of $\pop{A}{\dataset}\times\pop{B}{\dataset}$.
   If there are no violations, that is: $\viol{u}{\dataset} = \emptyset{}$, then we say that $u$ is satisfied.

>>>>

\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be updated regularly to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly updates of existing systems.
   Schema changes cannot always be avoided when updating software, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   The risk and effort of such data migrations explains why development teams try to avoid schema changes at all costs.

   Data migration for other purposes than schema change has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers can and will avoid schema changes and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
   Another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014}, defines a migration as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   This does not cater for schema changes, however.
   Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
   If the schema is not available at compile time, the work will have to be done at runtime.
   This requires versioned storage of production data and an overhead in performance.
   This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
   The prototypical use case for that is to release updates of information systems in production,
   where the semantic integrity of data must be preserved across schema changes.
   Another use case is application integration for multiple dispersed data sources with explicit schemas.

   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, invariants that change yield extra work for software engineers to write the transform code,
   for which ETL tools typically provide little support.
   The ubiquitous pressure to decrease deploy times and increase release frequencies calls for further automation of SCDMs.
   Hence, any downtime for the sake of migration becomes less and less acceptable.
   That is why our research aims for zero downtime.

   Another practical problem is that of data quality.
   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean it up.
   Some of that work must be done before the migration; some can wait till after the migration.
   We can capture part of the data quality problem as a requirement to satisfy semantic constraints.
   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
   a registration of valid addresses can be checked automatically.
   In a formalism like Ampersand, which allows us to express such constraints, we can add constraints for data quality to the schema.
   This allows us to signal the data pollution at runtime.
   Some forms of data pollution cannot be detected in this way, however.
   An example is when a person has deliberately specified a false name without violating any constraint in the system.

@@

\section{Analysis}
\label{sct:Analysis}
   This section analyzes information systems, so we can formally define the migration procedure from one system to another.
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
   The data in the system constitutes the collective memory of the business,
   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
   %This paper uses semantic constraints to represent this meaning.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
   \end{center}
\caption{Anatomy of an information system}
\label{fig:pre-migration}
\end{figure}

   Actors (both users and computers) are changing the data in a system continually.
   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
   Events that the system detects may cause the state to change.
   To keep our theory technology independent, we assume data sets to contain triples.
   This makes our theory valid for any kind of database that triples can represent,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.

   We assume that the semantics of the data is ensured by means of constraints.
   In figure~\ref{fig:pre-migration} they are depicted in the schema s rules.
   We distinguish two types of constraint:
\begin{enumerate}
\item Invariant\\
   An \define{invariant} is a constraint that is always true in a system.
   It serves to constrain the data set at runtime, to ensure its semantic integrity.
   % Violation of an invariant means it has to be restored without the outside world noticing.
   % This requires either human intervention or an automated procedure.
   % Once it is restored, the system commits to the new state, which satisfies the constraint.
   % If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
   % In both cases, the invariant remains true.

   A special case of an invariant is the \define{enforcement rule}.
   This corresponds to the classical database transaction.
   As soon as an enforcement rule is violated, the system restores it automatically, without human intervention,
   so the outside world will never notice that it was violated.
\item Business constraint\\
   A \define{business constraint} is a constraint that can be violated temporarily until a user restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   We consider business constraints not to be invariants because users can violate them.
   They are not true all the time.
\end{enumerate}

\subsection{Zero downtime}
   To make the case for zero downtime, consider this problem:
   Suppose we have an invariant, $u$, in the {\em desired system}, which is absent in the {\em existing system}.
   And suppose the data in the existing system does not satisfy $u$.
   When spinning up the desired system on the data set of the existing system, $u$ must be satisfied.
   If $u$ is an enforcement rule, the desired system will restore it automatically.
   But otherwise, the system will refuse to spin up.
   In many cases, clearing out all violations before spinning up the desired system might take too long, while new violations may arise in the meantime.
   So, the problem lies with invariants that are not enforced automatically.
   To solve this problem, we must implement such invariants initially as a business constraint,
   letting users restore invariance while the new system is running.
   For this purpose, we must define an intermediate system, the \define{migration system},
   which implements the invariants of the desired system as business constraints, unless they are enforcement rules.
   This migration system must also ensure that every violation that is fixed will not return as a violation.
   In this way, the business constraint eventually becomes an invariant, satisfying the specification of the desired system.
   Since the number of violations is finite, the business can resolve these violations in finite time.
   New violations that occur are under control of the desired system,
   so the business experiences the migration system as the desired system.
   In this way, the migration system bridges the gap and users get a zero downtime SCDM.

@@

@@

   Before the migration, the data in the existing system satisfies its invariants (by definition).
   When the migration system compiles correctly, it satisfies its own invariants,
   so it can be deployed in production.
   That is the MoT.
   The task of the migration system is to satisfy the invariants of the desired system
   so that it can be deployed in production.
   For some of these invariants, enforcement rules can satisfy them automatically.
   There may be invariants, however, that require human interaction to resolve.
   Once all invariants of the desired system are satisfied, the migration is ready for the MoC.
   So, the schema of the migration system must contain the rules to satisfy all invariants of the desired system.

@@

\subsection{Violations}
\label{sct:Violations}
   To satisfy a constraint $u$ means that it is true for all triples in the data set.
   Let us assume that we can represent each violation of a constraint $u$ as a pair
   and let the function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produce these violations.
   Note that $\viol{u}{\dataset}=\emptyset$ means that $\dataset$ satisfies $u$.
   Now we can represent an invariant $u$ as a triple $\triple{\violC{u}}{A}{B}$.
   It must be well-typed, just like a relation:
\begin{equation}
   \pair{a}{b}\in\viol{\triple{\violC{u}}{A}{B}}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
\end{equation}
   We represent an enforcement rule $u$ as a pair $\pair{\declare{n}{A}{B}}{\violC{u}}$,
   where relation $\declare{n}{A}{B}$ is used to restore invariance of the enforcement rule.
   The enforcement rule must be well-typed:
\begin{equation}
   \pair{a}{b}\in\viol{\pair{\declare{n}{A}{B}}{\violC{u}}}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
\end{equation}
   Besides, an enforcement rule must be automatically restorable by adding the violations to relation $\declare{n}{A}{B}$:
\begin{equation}
   \viol{u}{\dataset\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}=\emptyset
\end{equation}
   The language Ampersand has more types of enforcement rules than just this one,
   but to define this one only is sufficient for this paper.

\subsection{Schemas}
\label{sct:Schemas}
   Schemas serve to capture the semantics of a system~\cite{Spivak2012}.
   They define concepts, relations, and invariants.
   We assume that a software engineer defines a schema on design time, and a compiler checks whether the semantics are consistent.
   Error messages from the compiler are prohibitive for generating code,
   which prevents a substantial class of mistakes to ever reach end users.

@@

   Every invariant is an element of an infinite set called $\Rules$.
   In this paper, variables $u$ and $v$ represent rules.
   In Ampersand, rules are well-typed, just like relations:
   Every violation $\pair{a}{b}$ of a invariant $\triple{\violC{u}}{A}{B}$ must satisfy $a\instance A$ and $b\instance B$.
   % For every data set $\dataset$, the set $\viol{u}{\dataset}$ is a subset of $\pop{A}{\dataset}\times\pop{B}{\dataset}$.
   If there are no violations, that is: $\viol{u}{\dataset} = \emptyset{}$, then we say that $u$ is satisfied.
